---
  title: "Forecast Metro train delays in and around NYC"
author: "Nupur Baghel, Shivani Rai"
date: "December 15th, 2020"
output: 
  html_document:
  toc: true
toc_float: true
code_folding: "hide"
code_download: true
---
  
  
  ```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r setup_13, message=FALSE, include=FALSE}
library(tidyverse)
library(sf)
library(lubridate)
library(tigris)
library(tidycensus)
library(viridis)
library(riem)
library(gridExtra)
library(knitr)
library(kableExtra)
library(RSocrata)
library(caret)
library(osmdata)
library(gifski)
library(gganimate)
library(mapview)

mapTheme <- function(base_size = 12) {
  theme(
    text = element_text( color = "black"),
    plot.title = element_text(size = 14,colour = "black"),
    plot.subtitle=element_text(face="italic"),
    plot.caption=element_text(hjust=0),
    axis.ticks = element_blank(),
    panel.background = element_blank(),axis.title = element_blank(),
    axis.text = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill=NA, size=2)
  )
}

plotTheme <- theme(
  plot.title =element_text(size=12),
  plot.subtitle = element_text(size=8),
  plot.caption = element_text(size = 6),
  axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
  axis.text.y = element_text(size = 10),
  axis.title.y = element_text(size = 10),
  # Set the entire chart region to blank
  panel.background=element_blank(),
  plot.background=element_blank(),
  #panel.border=element_rect(colour="#F0F0F0"),
  # Format the grid
  panel.grid.major=element_line(colour="#D0D0D0",size=.2),
  axis.ticks=element_blank())

mapTheme <- theme(plot.title =element_text(size=12),
                  plot.subtitle = element_text(size=8),
                  plot.caption = element_text(size = 6),
                  axis.line=element_blank(),
                  axis.text.x=element_blank(),
                  axis.text.y=element_blank(),
                  axis.ticks=element_blank(),
                  axis.title.x=element_blank(),
                  axis.title.y=element_blank(),
                  panel.background=element_blank(),
                  panel.border=element_blank(),
                  panel.grid.major=element_line(colour = 'transparent'),
                  panel.grid.minor=element_blank(),
                  legend.direction = "vertical", 
                  legend.position = "right",
                  plot.margin = margin(1, 1, 1, 1, 'cm'),
                  legend.key.height = unit(1, "cm"), legend.key.width = unit(0.2, "cm"))

palette5 <- c("#eff3ff","#bdd7e7","#6baed6","#3182bd","#08519c")
palette4 <- c("#D2FBD4","#92BCAB","#527D82","#123F5A")
palette3 <- c("#D2FBD4","#92BCAB","#527D82")
palette2 <- c("#6baed6","#08519c")

palette55 <- c("#25CB10", "#5AB60C", "#8FA108",   "#C48C04", "#FA7800")
qBr <- function(df, variable, rnd) {
  if (missing(rnd)) {
    as.character(quantile(round(df[[variable]],0),
                          c(.01,.2,.4,.6,.8), na.rm=T))
  } else if (rnd == FALSE | rnd == F) {
    as.character(formatC(quantile(df[[variable]]), digits = 3),
                 c(.01,.2,.4,.6,.8), na.rm=T)
  }
}
q5 <- function(variable) {as.factor(ntile(variable, 5))}
```

```{r install_census_API_key, warning = FALSE, include=FALSE, eval = TRUE}
# Install Census API Key
census_api_key("e79f3706b6d61249968c6ce88794f6f556e5bf3d", overwrite = TRUE)
```

# Import Data

## Kaggle Train Delay dataset

```{r}
dat = read.csv('archive/2018_03.csv')

#Convert all lines to lower case
dat$line = tolower(dat$line)
dat$scheduled_time = as_date(dat$scheduled_time)
dat$actual_time = as_date(dat$actual_time)
```

## New Jersey Rail Lines and Rail Stations dataset

```{r}
njlines <- st_read('Passenger_Railroad_Lines_in_NJ-shp/Passenger_Railroad_Lines_in_NJ.shp') %>%
  st_transform('EPSG:2236')
njstations <- st_read('Railroad_Stations_in_NJ-shp/Railroad_Stations_in_NJ.shp') %>%
  st_transform('EPSG:2236')
amtrak <- st_read('Amtrak_Stations-shp/Amtrak_Stations.shp') %>%
  st_transform('EPSG:2236')
```


```{r}
hist(dat$delay_minutes)
```

```{r message=FALSE, warning=FALSE}
subset_amtrak <- st_crop(amtrak, st_bbox(njstations))

ggplot() +
  geom_sf(data = subset_amtrak, color = "black") +
  geom_sf(data = njlines, color = "blue") +
  mapTheme
```

```{r}
# dat - nj (geometry)
# atl. city line - ATLANTIC CITY RAIL LINE
# bergen co. line - BERGEN COUNTY LINE, HUDSON BERGEN LIGHT RAIL
# main line - MAIN LINE
# montclair-boonton - MONTCLAIR BOONTON LINE
# morristown line - MORRIS & ESSEX
# no jersey coast - NORTH JERSEY COAST LINE
# northeast corrdr- NORTHEAST CORRIDOR LINE
# pascack valley - PASCACK VALLEY LINE
# raritan valley - RARITAN VALLEY LINE

# dat - Amtrak

# [1] "regional"                  "amtrak"                   
# [3] "crescent"                  "acela express"            
# [5] "keystone"                  "pennsylvanian"            
# [7] "carolinian"                "silver star  -r"          
# [9] "silver meteor-r"           "vermonter    -r"          
# [11] "amtrak regional"           "palmetto"                 
# [13] "cardinal"                  "bergen co. line "         
# [15] "silver service / palmetto"

# Missing - 6 (gladstone branch), 14 (princeton shuttle)

#amtrak_type_line <- dat %>% filter (type == 'Amtrak')
#dat_amtrak <- subset(dat, type == 'Amtrak')
#dat_amtrak_from <- dat_amtrak %>%
#left_join(amtrak, by = c('from' = 'CITY2'))
#dat_amtrak_to <- dat_amtrak %>%
#left_join(amtrak, by = c('to' = 'CITY2'))


dat <- dat %>%
  mutate(new_line_name_geo = case_when(
    line == 'atl. city line' ~ 'ATLANTIC CITY RAIL LINE',
    line == 'bergen co. line' ~ 'BERGEN COUNTY LINE',
    line == 'main line' ~ 'MAIN LINE',
    line == 'montclair-boonton' ~ 'MONTCLAIR BOONTON LINE',
    line == 'morristown line' ~ 'MORRIS & ESSEX',
    line == 'no jersey coast' ~ 'NORTH JERSEY COAST LINE',
    line == 'northeast corrdr' ~ 'NORTHEAST CORRIDOR LINE',
    line == 'pascack valley' ~ 'PASCACK VALLEY LINE',
    line == 'raritan valley' ~ 'RARITAN VALLEY LINE',
  ))
dat_nj <- subset(dat, type != 'Amtrak')

dat_nj_line <- dat_nj %>%
  left_join(njlines, by = c('new_line_name_geo' = 'RAIL_LINE'))

dat_njstation <- dat_nj %>%
  left_join(njstations, by = c('from' = 'STATION'))

dat_njstation_to <- dat_nj %>%
  left_join(njstations, by = c('to' = 'STATION'))

nas <- dat_njstation %>%
  filter (is.na(COUNTY))
```

## Census Data

Collecting census data for States - New Jersey, New York and Pennsylvania

```{r get_census, message=FALSE, warning=FALSE, cache=TRUE}
census <- 
  get_acs(geography = "tract", 
          variables = c("B01003_001", "B19013_001", 
                        "B02001_002", "B08013_001",
                        "B08012_001", "B08301_001", 
                        "B08301_010", "B01002_001"), 
          year = 2018, 
          state = c("NJ",'NY','PA'),
          geometry = TRUE, 
          output = "wide") %>%
  rename(Total_Pop =  B01003_001E,
         Med_Inc = B19013_001E,
         Med_Age = B01002_001E,
         White_Pop = B02001_002E,
         Travel_Time = B08013_001E,
         Num_Commuters = B08012_001E,
         Means_of_Transport = B08301_001E,
         Total_Public_Trans = B08301_010E) %>%
  select(Total_Pop, Med_Inc, White_Pop, Travel_Time,
         Means_of_Transport, Total_Public_Trans,
         Med_Age,
         GEOID, geometry) %>%
  mutate(Percent_White = White_Pop / Total_Pop,
         Mean_Commute_Time = Travel_Time / Total_Public_Trans,
         Percent_Taking_Public_Trans = Total_Public_Trans / Means_of_Transport) %>% 
  st_transform('EPSG:2236')
```
```{r}
subset_census <- census[st_crop(census, st_bbox(njstations)),]

mapview(st_bbox(census)) +
  mapview(njstations, cex = 2) +
  mapview(st_bbox(subset_census), col.regions = "orange")

```


```{r}
ggplot() +
  geom_sf(data = subset_census) +
  geom_sf(data = njstations %>% st_transform('EPSG:2236'), color = "blue") +
  mapTheme
```


## Weather Data

Collecting weather data for Trenton for the same duration as our original data set (March 2018). We make an assumption that overall weather patterns will remain similar throughout the states. For a more granular study it would make sense to include other location's weather data as well.

```{r import_weather, message = FALSE, warning = FALSE, cache = TRUE}
#Every hour
weather.Panel_hourly <- 
  riem_measures(station = "TTN", date_start = "2018-03-01", date_end = "2018-03-31") %>%
  dplyr::select(valid, tmpf, p01i, sknt)%>%
  replace(is.na(.), 0) %>%
    mutate(interval60 = ymd_h(substr(valid,1,13))) %>%
    mutate(week = week(interval60),
           dotw = wday(interval60, label=TRUE)) %>%
    group_by(interval60) %>%
    summarize(Temperature = max(tmpf),
              Precipitation = sum(p01i),
              Wind_Speed = max(sknt)) %>%
    mutate(Temperature = ifelse(Temperature == 0, 42, Temperature))
glimpse(weather.Panel_hourly)
#Every day
weather.Panel_daily <- 
  riem_measures(station = "TTN", date_start = "2018-03-01", date_end = "2018-03-31") %>%
  dplyr::select(valid, tmpf, p01i, sknt)%>%
  replace(is.na(.), 0) %>%
    mutate(interval60 = ymd_h(substr(valid,1,13))) %>%
    mutate(week = week(interval60),
           dotw = wday(interval60, label=TRUE)) %>%
    group_by(date(interval60)) %>%
    summarize(Temperature = max(tmpf),
              Precipitation = sum(p01i),
              Wind_Speed = max(sknt)) %>%
    mutate(Temperature = ifelse(Temperature == 0, 42, Temperature))
glimpse(weather.Panel_daily)
```

## Time-space Panel + Weather 

```{r}
dat_njstation$date <- as.Date(dat_njstation$date)
dat_njstation <- 
  dat_njstation %>%
  left_join(weather.Panel_daily, by = c('date' = 'date(interval60)'))
```

## Exploratory Data Analysis

### Weather Panel 

```{r}
grid.arrange(
  ggplot(weather.Panel_hourly, aes(interval60,Precipitation)) + geom_line() + 
  labs(title="Precipitation", x="Hour", y="Precipitation") + plotTheme,
  ggplot(weather.Panel_hourly, aes(interval60,Wind_Speed)) + geom_line() + 
    labs(title="Wind Speed", x="Hour", y="Wind Speed") + plotTheme,
  ggplot(weather.Panel_hourly, aes(interval60,Temperature)) + geom_line() + 
    labs(title="Temperature", x="Hour", y="Temperature") + plotTheme,
  top="Weather Data - Trenton TTN - March, 2018")
```

### Delay vs Precipitation (Daily)

```{r}
dat_njstation %>%
  group_by(date) %>% 
  summarize(delay_minutes = mean(delay_minutes, na.rm=T),
            Precipitation = first(Precipitation)) %>%
  mutate(isPrecip = ifelse(Precipitation > 0,"Rain/Snow", "None")) %>%
  group_by(isPrecip) %>%
  summarize(Mean_delay = mean(delay_minutes, na.rm=T)) %>%
    ggplot(aes(isPrecip, Mean_delay)) + geom_bar(stat = "identity") +
      labs(title="Does precipitation influence delay?",
           x="Precipitation", y="Mean Delay (Minutes)") +
      plotTheme
```

### Delay vs Windspeed (Daily)


### Delay vs Temperature (Daily)


### Distribution of Total delay for different trains

```{r message=FALSE, warning=FALSE}
delay_train <- dat_njstation %>%
  mutate(date = date(scheduled_time)) %>%
  select(train_id, delay_minutes, date) %>%
  group_by(train_id, date) %>%
  summarize(total_delay = sum(delay_minutes))

ggplot(delay_train, aes(total_delay)) +
  geom_histogram()
```

### Distribution for Time of Day

Seems like we dont have time of the day - lol

### Distribution across Time and Space



### Reported Delays per Day

```{r message=FALSE, warning=FALSE}
dat_njstation <- dat_njstation %>%
  mutate(delay_Y.N = ifelse(delay_minutes == 0.00, 0, 1))
no_delay_per_day <- dat_njstation %>%
  mutate(date = date(scheduled_time)) %>%
  select(delay_Y.N, date) %>%
  group_by(date) %>%
  summarize(total_delay = sum(delay_Y.N))

ggplot(data = no_delay_per_day, aes(x = date, y = total_delay)) + geom_line() + geom_point() +
labs(title="Total Number of Delays", x="Date", y="Delays") + plotTheme

```

### Total Delays in Minutes per Day

```{r message=FALSE, warning=FALSE}
delay_per_day <- dat_njstation %>%
  mutate(date = date(scheduled_time)) %>%
  select(delay_minutes, date) %>%
  group_by(date) %>%
  summarize(total_delay = sum(delay_minutes),
            average_delay = mean(delay_minutes))

grid.arrange(
ggplot(data = delay_per_day, aes(x = date, y = total_delay)) + geom_line() + geom_point() +
labs(title="Total Delay", x="Date", y="Delay (Minutes)") + plotTheme,
ggplot(data = delay_per_day, aes(x = date, y = average_delay)) + geom_line() + geom_point() +
labs(title="Average Delay", x="Date", y="Delay (Minutes)") + plotTheme,
top="Total and Mean Delay - 1st to 31st March, 2018")
  
```

### Delays at stations - for all lines 

```{r}
dat_njstation <- dat_njstation %>% 
  mutate(Geom = gsub('c[()°]', '', geometry)) %>% 
  separate(col = Geom, into = c('Latitude', 'Long_delete'), sep = '\\,')
dat_njstation <- extract(dat_njstation, geometry, into = c('Lat_delete', 'Longitude'), '\\((.*),(.*)\\)', conv = T)


dat_njstation %>% 
  group_by(from) %>%
  summarize(mean_delay = mean(delay_minutes, na.rm = TRUE))%>%
ggplot(.)+
  geom_sf(data = census %>% st_transform('EPSG:2236'), color = "grey", fill = "transparent")+
  geom_point(aes(x = 'Latitude', y= 'Longitude', color = mean_delay), 
             fill = "transparent", alpha = 0.8)+
  scale_colour_viridis(direction = -1,
  discrete = FALSE, option = "D")+
  labs(title="Mean Abs Error, Test Set, Model 5")+
  mapTheme
```


+
123.0232

## Feature Engineering

### Distance between Stations

```{r}

```

### Correlation plot for Continuous numeric features

## Training Model

### Split Data into Train and Test
### 


There can be multiple factors causing delay. It is hard for us to be able to account for each. [NJ Transhit](https://njtranshit.com/excuses) provides a list of common and new excuses provided by rail authorities for delays. 